The analogue method for precipitation prediction: finding better analogue situations at a sub-daily time step

### 2017-03-27 Moving temporal window - HESS/template.tex
HESS paper under work.

@@

	The AMs were applied to both ERA-20C and MERRA-2 datasets. Most results are presented for the ERA-20C dataset, as it allows testing a 3h-MTW, but would lead to similar conclusions with MERRA-2. The impact of an MTW was assessed on two different basis:
	\begin{itemize}
		\item Original parameters: The AMs were initially calibrated for each station based on the classic 24-hr approach. Then, the MTW was introduced into the AMs, but the parameters (spatial windows and number of analogues) were not reassessed. The analogy is the same and thus initial analogue dates should be found in the new distribution, along with new dates provided by the MTW.
		\item Recalibrated: The parameters of the AMs were reassessed after the introduction of the MTW at different time steps. This allowed to improve the prediction skill of the AMs by adapting to the new information available through the MTW. The characterization of the analogy might thus change (different spatial windows and number of analogues).
	\end{itemize}

@@

	\subsubsection{Changes in the analogy criteria distributions}
	\label{sec:changes_distributions}

	\subsubsection{Changes in the atmospheric circulation analogy}
	\label{sec:perf_2Z}

	When searching for analogues in the first level of analogy, such as on the geopotential heights in the 2Z method (Sect. \ref{sec:analog_method}), there are 4 times as many candidates (even though not fully independent) with the MTW than before (Sect. \ref{sec:mtw}). Figure \ref{fig:changes_S1_analogs} shows the changes in the distributions of the analogy criterion (S1) for the $1^{st}$, $5^{th}$, $20^{th}$, and $40^{th}$ analogue rank at the Ulrichen station over the whole calibration period (1961\textendash 2008 \textendash\ the full period could be used here, as precipitation was not considered at this stage), due to introduction of the MTW. The shape of the distributions of the conventional approach and the MTW were found to be similar, but the values of the analogy criterion were reduced (shifted to the left) and were, therefore, better. For analogues with higher ranks (e.g. $20^{th}$ or $40^{th}$), the difference between the two distributions was larger than for the first ranks, which means that the improvement increased with the rank of the analogues.

	The improvements of the analogy with the rank of the analogues are summarized in Fig. \ref{fig:changes_S1}, which shows (top) quantiles of the S1 criterion for the conventional method and the MTW, and (bottom) quantiles of the relative reduction (meaning improvement) due to the MTW. This confirms that all quantiles were similarly reduced (S1 distributions keep their shape), and that this improvement was constantly increasing from the first to the last analogue (Fig. \ref{fig:changes_S1} bottom).

	For the first level of analogy, the improvement of the S1 criterion (Fig. \ref{fig:changes_S1} bottom) started approximately at 5~\% for the first analogue and reached more than 10~\% for the last (40$^{th}$) (Sect. \ref{sec:perf_2Z}). This increase in improvement with the analogue rank can be explained by the accumulation of better analogues in the selection, with new better situations pushing previous analogues to higher ranks. The curve representing the minimum improvement is mostly superior to zero, meaning that the criteria have been improved on most analogue ranks for every day of the calibration period. All other stations had a similar improvement in the S1 criterion, both in terms of the distribution shape and amplitude. The introduction of the MTW allows finding better analogue situations in the first level of analogy, resulting in a selection of days with better similarity in atmospheric circulation.

	When adding a second level of analogy with moisture variables, the criteria values (RMSE) were not improved by the MTW (Sect. \ref{sec:perf_2Z-2MI}), as the number of candidates was not higher (still the $N_{1}$ days selected in the first level of analogy). However, this result of a globally null improvement of the RMSE values does not mean that the 2Z-2MI method cannot be improved by the MTW. It means that after the selection of the analogue situations for the synoptic circulation, the new candidate dates did not provide better analogues in terms of moisture. However, the selected dates have changed in the first level of analogy and also in the final selection, and thus the distributions of precipitation values were different, which did improve the prediction skill (Sect. \ref{sec:influence_scores}).

	\subsubsection{Changes per precipitation class}
	\label{sec:influence_precip}

	The impact of the MTW on the analogy criterion has been analysed per precipitation classes (for the target day). The results are summarised in Fig. \ref{fig:changes_S1_precip_threshold} by the median reduction of S1 for the days with precipitation (organized into classes) between two thresholds. With the number of cases per class being reduced, the curves are not as smooth as in previous analyses. It is nevertheless clear that the improvements were larger for days with higher precipitation. Once again, the results for all the other stations were similar.

	\subsubsection{Changes per season}
	\label{sec:seasonal_effect}

	Atmospheric dynamics vary greatly from one season to another, reflecting on the performance of the AM that is generally lower between June and August \citep{Bliefernicht2010}. The effect of the MTW on the S1 criterion per season is presented in Fig. \ref{fig:changes_S1_seasons}. Differences between the seasons were substantial, with greater improvements for winter than summer.

	Section \ref{sec:seasonal_effect} and Fig. \ref{fig:changes_S1_seasons} revealed a difference in the improvement of the S1 criterion according to the season, with greater improvements for winter than summer. One hypothesis is that the diurnal effects of the summer months have an influence on the atmospheric circulation at least in the lower layers. This effect is in phase with the daily cycle, and good analogues are essentially found for the same hours.

	An analysis of the selected hours for the geopotential height predictor seems to confirm this assumption (Fig. \ref{fig:hours_selection_per_season}). It was found that the new choice of the temporal window in winter, when using the MTW approach, is well balanced between the four options. This means a change of 75~\% of the analogues selected compared to the conventional approach. On the contrary, the days during the summer months had a preference for the initial temporal window (Z500 24~h \& Z1000 12~h), likely due to more pronounced diurnal effects, which reduced the potential for improvement of the criteria. The other seasons were between these two extremes, which is consistent with their respective improvements. This seasonal effect was observed for each station in a very similar way and even with slightly larger amplitude than for Ulrichen.

	\subsubsection{Changes in the moisture analogy}
	\label{sec:perf_2Z-2MI}

	When adding the second level of analogy of the 2Z-2MI method (Table \ref{table:method_2Z-2MI}), the number of candidate situations did not increase, as they were conditioned by the $N_{1}$ previously selected analogues, but their dates had changed. In contrast to the AM on the atmospheric circulation only, both a reduction or an increase of the RMSE analogy criterion values were possible with the MTW compared to the static approach. Indeed, Fig. \ref{fig:changes_RMSE} shows an almost insignificant improvement of the RMSE values. Unlike the first level of analogy, the relative changes of the RMSE values were distributed relatively symmetrically around zero, with improvements and losses of the same amplitude.

	\subsection{Impact of the MTW on prediction skill}
	\label{sec:influence_scores}

	The new performance scores (CRPS) of the precipitation prediction are provided in Fig. \ref{fig:plots_CRPS_2Z} and \ref{fig:plots_CRPS_2Z-2MI} for the 2Z and 2Z-2MI methods, respectively, for the reduced archive (1982\textendash 2007).

	Both the analogy criteria (Sect. \ref{sec:influence_precip}) and the performance scores (Sect. \ref{sec:improvement_CRPS_precip_threshold}) were improved to a greater extent for days with heavier precipitation. This is likely due to the fact that higher precipitation events are a consequence of atmospheric conditions with greater dynamics, such as weather disturbances, which have a well-marked temporal evolution. Indeed, the position of the driving elements, such as the low pressure centres and the fronts, change significantly during a day. These situations are less numerous than anticyclonic situations, which makes it less likely to find very good analogues at the same time of the day. We can, therefore, expect to improve these situations with greater dynamics more significantly when introducing a MTW, as better matches to the target situation may be found. In contrast, days with low dynamics in the atmospheric circulation, such as anticyclonic situations, will not be radically improved by the introduction of the MTW.

@@

	\subsubsection{Improvement per precipitation classes}
	\label{sec:improvement_CRPS_precip_threshold}

	The S1 criterion was previously found to show greater improvement for higher precipitation values (Sect. \ref{sec:influence_precip}). The changes in terms of prediction skill were also assessed for the precipitation classes. Fig. \ref{fig:changes_CRPS_precip_threshold} synthesizes these differences for the Ulrichen station, with the other stations having the same behaviour. The performance score was improved for days with higher precipitation after the introduction of the MTW. In contrast, for non-rainy days and small precipitation values, the performance scores were not improved.

	\subsubsection{Recalibration of AM parameters}
	\label{sec:recalibration}

	The previous assessment of the performance improvement was established with the original parameters optimized with a fixed time window. However, one can assume that the introduction of the MTW might change the optimum value of some parameters. The calibration (see Sect. \ref{sec:calibration}) should then be reprocessed.

	After recalibrating, changes in the AM parameters could be observed for both the 2Z and 2Z-2MI methods. Among these, the zonal extent of the spatial windows of the circulation analogy decreased slightly (not shown). More significantly, the optimal number of analogues $N_{1}$ and $N_{2}$ increased after introducing the MTW by a considerable magnitude (Fig. \ref{fig:analog_nb}): 25~\% to 83~\% for the 2Z method and 20~\% to 67~\% for the final selection of the 2Z-2MI method. The number of analogues $N_{1}$ of the first analogy level of the 2Z-2MI method reached three times its previous value for the Visp station ($N_{1}$~=~135 instead of 45; Fig. \ref{fig:analog_nb}).

	The values of the CRPS scores for both methods (Fig. \ref{fig:plots_CRPS_2Z} and \ref{fig:plots_CRPS_2Z-2MI}) have significantly increased after recalibration. When analysing the change in performance per precipitation class, the results (not shown) were very similar to the observations in Sect. \ref{sec:improvement_CRPS_precip_threshold}, with a slight performance increase for small precipitation values that can be observed at the expense of higher amounts, due to the higher number of analogues.

	\subsection{Consequences of the archive reduction}
	\label{sec:archive_reduction}

@@

	\subsection{On the increase in the number of analogues}

	After the recalibration of the AMs with the MTW, the optimal analogue numbers were significantly higher than in the original methods (see Sect. \ref{sec:recalibration} and Fig. \ref{fig:analog_nb}). As shown in Fig. \ref{fig:changes_S1}, the improvement of the S1 criterion grew along with the rank of the analogue, which shows an accumulation of better analogue situations in the distributions. It seems profitable to widen the selection of analogues in order to also keep some whose rank has increased, as they appear to be relevant to the prediction of the precipitation values. The number of good analogues was globally increased.

	\subsection{Why not just make 6-hourly predictions?}

	One can question the interest of using moving daily totals when 6-hourly precipitation series can be predicted instead. However, the 6-hourly time series generated by the AM might not represent accurately the intra-daily precipitation distribution (results not shown). In addition, sometimes a resolution finer than the daily time step is not needed and another disaggregation technique may be used afterwards. Finally, when using a reconstructed precipitation archive, the errors in intra-daily precipitation distributions have a smaller impact than 24h daily totals.

	\subsection{Reconstruction of a longer precipitation archive}
	\label{sec:reconstruction}

>>>>

	%TODO: improve
	One can question the interest of using moving daily totals when 6-hourly precipitation series can be predicted instead. However, the 6-hourly time series generated by the AM might not represent accurately the intra-daily precipitation distribution (results not shown). In addition, sometimes a resolution finer than the daily time step is not needed and another disaggregation technique may be used afterwards. Finally, when using a reconstructed precipitation archive, the errors in intra-daily precipitation distributions have a smaller impact than 24h daily totals.

@@

	The AMs were applied to both ERA-20C and MERRA-2 datasets. Most results are presented for the ERA-20C dataset, as it allows testing a 3h-MTW, but would lead to similar conclusions with MERRA-2. The impact of an MTW was assessed on two different basis:
	\begin{itemize}
		\item Original parameters: The AMs were initially calibrated for each station based on the classic 24-hr approach. Then, the MTW was introduced into the AMs, but the parameters (spatial windows and number of analogues) were not reassessed. The analogy is the same and thus initial analogue dates should be found in the new distribution, along with new dates provided by the MTW.
		\item Recalibrated: The parameters of the AMs were reassessed after the introduction of the MTW at different time steps. Indeed, one can assume that the introduction of the MTW might change the optimum value of some parameters (different spatial windows and number of analogues). The calibration (see Sect. \ref{sec:calibration}) was then be reprocessed to adapt the AMs to the new information available through the MTW. The main difference was the increase of the optimal number of analogues, which is discussed in Sect. \ref{sec:increase_analognb}.
	\end{itemize}

@@

	First, the impact of the MTW on the selection of analogue dates is analyzed, independently of any precipitation data. This was done on the period 1982-2010, but could have been done on a longer period, as no sub-daily precipitation series is required at this stage. The results are presented for the Ulrichen station, but are very similar for all other stations. The original parameters were used here when assessing the impact of the MTW, no recalibration of the parameters took place.

	\subsubsection{Analogy of the atmospheric circulation}
	\label{sec:changes_analogy_circul}

	When searching for analogues in the first level of analogy, such as on the geopotential heights in the 2Z method (Sect. \ref{sec:analog_method}), there are up to eight times as many candidates (even though not fully independent) with the MTW than before (Sect. \ref{sec:mtw}). Figure \ref{fig:changes_S1_analogs} shows the changes in the distributions, over the whole period, of the analogy criterion (S1) for the $1^{st}$, $15^{th}$, and $30^{th}$ analogue rank, due to introduction of an MTW with a 12h, 6h, or 3h time step. The shape of the distributions of the conventional approach and the MTWs were found to be similar, but the values of the analogy criterion were gradually reduced (shifted to the left) with smaller time steps MTWs and were, therefore, better. The circulation analogy seems to improve regularly with the doubling of the MTW time steps. For analogues with higher ranks (e.g. $15^{th}$ or $30^{th}$), the difference between the two distributions was larger than for the first rank, which means that the improvement increased with the rank of the analogues.

	The improvements of the analogy with the rank of the analogues are summarized in Fig. \ref{fig:changes_S1}, which shows (top) quantiles of the S1 criterion for the conventional method and the MTWs at different time steps, and (bottom) quantiles of the relative reduction (meaning improvement) due to the MTWs. This confirms that all quantiles were similarly reduced (S1 distributions keep their shape), and that this improvement was constantly increasing from the first to the last analogue (Fig. \ref{fig:changes_S1} bottom). This increase in improvement with the analogue rank can be explained by the accumulation of better analogues in the selection, with new better situations pushing previous analogues to higher ranks. One can also see the regularity of the improvements brought by decreasing time steps. The introduction of the MTW allows finding better analogue situations in the first level of analogy, resulting in a selection of days with better similarity in atmospheric circulation. Moreover, this improvement is superior for MTWs with a smaller time step.

	The impact of the MTW on the analogy criterion has been analysed per precipitation classes (for the target day) for a 6h-MTW and a 3h-MTW. The results are summarised in Fig. \ref{fig:changes_S1_precip_threshold} by the median reduction of S1 for the days with precipitation (organized into classes) between two thresholds. With the number of cases per class being reduced, the curves are not as smooth as in previous analyses. It is nevertheless clear that the improvements were larger for days with higher precipitation, for both the 6h-MTW and the 3h-MTW.

	The effect of the MTW on the S1 criterion has been decomposed per season and is presented in Fig. \ref{fig:changes_S1_seasons}. Differences between the seasons are substantial, with greater improvements for winter than summer. An analysis of the selected hours for the geopotential height predictor was then performed for the MTW with different time steps (Fig. \ref{fig:hours_selection_per_season}). It was found that the new choice of the temporal window in winter, when using the MTW approach, is more balanced between the different hours of the day (more regular repartition), and this for all time steps. This means a change in selection of a greater part of the analogue dates compared to the conventional approach. On the contrary, the days during the summer months had a preference for the initial temporal window (Z500 24~h \& Z1000 12~h), likely due to more pronounced diurnal effects, which reduced the potential for improvement of the criteria. This effect is in phase with the daily cycle, and good analogues are essentially found for the same hours. When a 3h-MTW is used, the time step following the initial temporal window (so Z500 27~h \& Z1000 15~h) is almost as often selected for all seasons. However, it can be considered as relatively similar in terms of the solar influence. The other seasons were between these two extremes, which is consistent with their respective improvements.

	\subsubsection{Moisture analogy}
	\label{sec:changes_analogy_moisture}

	When adding the second level of analogy of the 2Z-2MI method (Table \ref{table:method_2Z-2MI}), the number of candidate situations did not increase, as they were conditioned by the $N_{1}$ previously selected analogues, but their dates had changed. In contrast to the AM on the atmospheric circulation only, both a reduction or an increase of the RMSE analogy criterion values were possible with the MTW compared to the static approach. The results of the second level of analogy are presented in Fig. \ref{fig:changes_RMSE}, and show no improvement of the RMSE values, whatever the MTW time step. Unlike the first level of analogy, the relative changes of the RMSE values were distributed relatively symmetrically around zero.

	The criteria values (RMSE) of the second level of analogy (moisture variables) were not improved by the MTW, as the number of candidates was not higher (still the $N_{1}$ days selected in the first level of analogy). However, this result of a globally null improvement of the RMSE values does not mean that the 2Z-2MI method cannot be improved by the MTW. It means that after the selection of the analogue situations for the synoptic circulation, the new candidate dates did not provide better analogues in terms of moisture. However, the selected dates have changed in the first level of analogy and also in the final selection, and thus the distributions of precipitation values were different.

	\subsection{Impact of the MTW on the prediction skill}
	\label{sec:influence_scores}

	The change in the performance score (CRPS) of the precipitation prediction are provided in Fig. \ref{fig:plots_CRPS_2Z} and \ref{fig:plots_CRPS_2Z-2MI} for the 2Z and 2Z-2MI methods, respectively. The AMs were here recalibrated on the CP, and assessed on the independent VP. The score was processed for each station and both reanalysis datasets. For ERA-20C, the 12h-MTW, 6h-MTW, and 3h-MTW could be assessed, while only the two first could be processed with the MERRA-2 dataset.

	The MTW improves the precipitation prediction, as at least from a 6h-MTW the CRPS decreases for all stations, both on the CP and the VP. The VP shows a greater variability, likely related to the short period it represents (6 yr). The prediction skill on the CP is almost always improved further by reducing the time step of the MTW, but not of the same magnitude. For ERA-20C, the magnitude of the improvement of 2Z on the CP is relatively reduced when using a 3h-MTW instead of a 6h-MTW, and the variablility on the VP increases. However, as it was shown in Fig. \ref{fig:changes_S1}, the distribution of analogue dates was improved relatively equally for every reduction of the MTW time step. A possible explanation could be that the AMs rely on the geopotential at 500~hPa, while the ERA-20C dataset was build by only assimilating surface observations. Thus, the timing of the atmospheric circulation at higher pressure levels might not be completely in phase with the real perturbation systems that are linked with the observed precipitations. The MTW with the MERRA-2 dataset and the 2Z method also shows a certain slope break after the 12h-MTW, that could not be explained, and that is less important for the 2Z-2MI method. However, the 12h-MTW shows an important variability for the 2Z-2MI method and the MERRA-2 dataset. 
	Based on these results, the use of ERA-20C seems to be a better choice in order to apply an MTW, at least for the studied region.

	The S1 criterion was previously found to show greater improvement for days with higher precipitation values (Sect. \ref{sec:influence_precip}). The changes in terms of prediction skill were also assessed for the precipitation classes. Fig. \ref{fig:changes_CRPS_precip_threshold} synthesizes these differences for the Ulrichen station, on the original parameters. The performance score was improved (reduced CRPS) for days with higher precipitation after the introduction of the MTW. In contrast, for non-rainy days and small precipitation values, the performance scores were not improved. The recalibration of the parameters tends to reduce the spread of the changes in CRPS, but with a slight loss in performance for the most rainy days.

	%TODO: on the CP or the VP ? -> test on VP before/after recalib

	\subsection{A better prediction of heavy precipitation}

	Both the analogy criteria (Sect. \ref{sec:changes_analogy_circul}) and the performance scores (Sect. \ref{sec:influence_scores}) were improved to a greater extent for days with heavier precipitation. This is likely due to the fact that higher precipitation events are a consequence of atmospheric conditions with greater dynamics, such as weather disturbances, which have a well-marked temporal evolution. Indeed, the position of the driving elements, such as the low pressure centres and the fronts, change significantly during a day. These situations are less numerous than anticyclonic situations, which makes it less likely to find very good analogues at the same time of the day. We can, therefore, expect to improve these situations with greater dynamics more significantly when introducing a MTW, as better matches to the target situation may be found. In contrast, days with low dynamics in the atmospheric circulation, such as anticyclonic situations, will not be radically improved by the introduction of the MTW.

@@

	\subsection{The relationship between the MTW time step and the number of analogues}
	\label{sec:increase_analognb}

	When recalibrating the AMs with MTWs of different time steps, the optimal number of analogues changed for both 2Z and 2Z-2MI methods. $N_{1}$ and $N_{2}$, from the first and the second level of analogy, tend to significantly increase with the reduction of the time step of the MTW (Fig. \ref{fig:analog_nb}). When using ERA-20C, all optimal numbers of analogues (for the different analogy levels) tend in average to double when using a 6h-MTW instead of the classic approach, and to even triple when using a 3h-MTW. These higher numbers of analogues were objectivel chosen by the calibration procedure (Sect. \ref{sec:calibration}) in order to increase the prediction skill of the methods.

	As shown in Fig. \ref{fig:changes_S1}, the improvement of the S1 criterion grows along with the rank of the analogue, which shows an accumulation of better analogue situations in the distributions. It seems profitable to widen the selection of analogues in order to also keep some whose rank has increased, as they appear to be relevant to the prediction of the precipitation values. The number of good analogues was globally increased.

	\subsection{Inflation of the precipitation archive}
	\label{sec:archive_reduction}

	%TODO: do calculations of the conequence of the archive reduction
	%TODO: rewrite this section

@@

	\subsection{Reconstruction of a long precipitation archive suitable for the MTW}
	\label{sec:reconstruction}
